Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2025-01-14 manuscript/manuscript.tex
minor edits

\subsection{Structural Issues}
\label{sec:structural}
Structural issues refers to how data is organized, formatted, 
or presented within a dataset. Structural issues can cause difficulty in 
 analyzing the data effectively, often necessitating extensive 
data cleaning. We encountered three structural issues.

@@

\subsection{Accuracy and Precision:}
\label{sec:precision}
The question of precision vs. accuracy arises with the \texttt{latitude} 
and \texttt{longitude} fields. as well as with the duplicative 
\texttt{location} field. These fields are expressed as
a 14-decimal number, e.g. \texttt{latitude} 40.86769186022511. 
Given that 1 degree of latitude at the equator is equal to 111.04 
kilometers, the ``1'' at the end of that decimal number represents 
approximately 1.1 nanometers (1/1,000,000,000 of a meter). (For 
reference a DNA molecule is approximately 2 nanometers in width.) This 
representation appears to be a classic case of extreme precision, albeit with limited 
accuracy. All the \texttt{latitude}, \texttt{longitude}, and \texttt{location}
fields display this characteristic.

@@

\section{Government Open Data Curation Recommendations}
\label{sec:recommendations}
Building upon  insights from our case study of the 
NYC 311 SR data, we propose a set of data 
curation principles tailored for government open datasets. 
These principles are designed to address the unique challenges 
 observed in the curation of such datasets, 
ensuring they remain reliable, consistent, and useful for 
public services, research, and other applications.

\subsection{Principles}
\paragraph{Principle 1: Establishing Cross-Organizational Consistency}
A key challenge in government released datasets is the 
harmonization of data fields across various Agencies, and 
maintaining that consistency over time. To address this, Agencies should adopt 
standardized naming conventions, value domains, usages, and 
formats across all datasets. Regular audits and data harmonization 
processes should be implemented to ensure consistency 
across long-term datasets, particularly as Agency structures 
and reporting standards evolve. For instance, ensuring that 
historical datasets reflect the same field naming conventions 
over a span of years can prevent inconsistent analyses. 
These standards should be promulgated by the Open Data 
governing authority, following agreements by the relevant parties.

\paragraph{Principle 2: Ensuring Data Entry Accuracy}
Data accuracy and validity are fundamental to the utility of 
open datasets. In the NYC 311 data, issues such as 
invalid zip codes and nonsensical date entries 
highlight the need for rigorous validation processes. Governments 
should establish clear protocols for data entry, ensuring that 
fields are populated with valid, legally acceptable values. To 
the maximum extent possible,the software should assist in 
ensuring data entries are accurate and valid. Some examples include:

@@

\paragraph{Principle 3: Optimizing Storage Efficiency}
Efficient storage and data representation are crucial for handling 
large government datasets. Our research found several duplicate and
near-duplicate fields in the 311 SR dataset. Eliminating 
redundant fields and encoding categorical variables can 
significantly reduce file sizes and improve the efficiency 
of data analysis. Similarly, categorical variables could be 
encoded in a standardized format to optimize storage and 
streamline analyses. Reducing file size not only saves 
storage space but also enhances the performance of queries, 
downloads,  and analysis processes.

\paragraph{Principle 4: Updating Data Dictionaries}
Clear and accurate data documentation is essential for the effective use 
of government datasets. Our study found several 
instances where the Data Dictionary did not accurately reflect the 
actual data. Domains of legal/allowable values should be referenced
in the Data Dictionary. Regular reviews and updates should 
occur whenever there are changes in data structure, data usage, or 
content. This practice enhances the ability of external users, 
including researchers and developers, to work 
effectively with the data.

\paragraph{Principle 5: Automating Quality Assurance Processes}
Automating data validation and quality assurance processes can 
significantly improve the reliability of government datasets. Certain
unusual trends can be identified, and irregularities highlighted. 
In the NYC 311 case, issues such as large spikes in SR 
creation and closure at exactly midnight (or noon) likely 
stem from a cron script process that distorts the 
accurate capture of SR durations. By implementing 
real-time and event-driven validation tools, governments 
can both prevent such errors from entering the dataset in the first 
place, as well as highlight anomalous trends. Automated 
systems can be designed to flag inconsistencies, incorrect values, 
illogical sequences, or illogical data entries. Automation can 
also facilitate faster detection and correction of errors, reducing 
the need for manual intervention and ensuring the data 
remains clean and reliable as it is continuously updated.

\paragraph{Principle 6: Addressing Data Transparency}
Transparency is a core tenet of open data, and governments must ensure 
that datasets are not only accurate but also easily accessible and 
readily understandable. This involves providing clear 
metadata and making datasets available in user-friendly 
formats. Additionally, governments should communicate transparently about 
data quality issues, such as notifying the public when errors in SR 
fields (e.g., erroneous \texttt{closed\_ date(s)} are discovered, including
an explanation about how such issues will be resolved. By 
ensuring transparency and keeping the public informed, 
governments foster trust and promote wider use of datasets 
for research, civic engagement, and innovation.

\subsection{Actionable Steps}
While these principles are broad, actionable steps are required to 
implement them effectively. Governments can begin by investing in 
real-time validation systems, regular updates to data dictionaries, 
and tools for encoding and standardizing data fields. This would include
systems that automate enforcement of data validation rules 
(e.g., flagging incorrect zip codes), significantly improving data 
quality. Collaboration with city agencies and stakeholders is crucial 
to ensure that data standards, formats, and metadata meet the needs 
of diverse users, including policymakers, researchers, and the 
general public. Regular communication between agencies and external 
users can help identify recurring issues and areas where improvements 
are needed. By facilitating regular communication and collaboration 
with various stakeholders, governments can ensure data remains relevant 
and actionable for public use.

Automated Quality Assurance techniques should be employed to 
monitor the quality of the data, as measured by agreed upon
standards, such as the \% of invalid zip codes. Such tools could
capture and display the quality of the data sets as presented
on appropriate dashboards. Deviations in quality would be
automatically highlighted, alerting appropriate corrective actions.

Engagement with data scientists and statisticians is key to improving 
data curation strategies. These experts can help design and refine 
curation processes, such as creating algorithms for automatic data 
cleaning, identifying inconsistent data entries, and ensuring the 
dataset(s) structure is maintained over time. Their expertise is also 
critical in developing protocols for validating incoming data, correcting 
errors in real-time, and ensuring consistency across multiple data sources. 
Feedback, such as the issues reported 
in this paper, benefit from a clear and accessible channel to the 311 SR data 
team and other relevant City Agencies. Establishing such feedback loops 
and other opportunities for public engagement, ensures that 
recommendations for improving data curation and validation 
are not only heard and discussed, but implemented. This feedback 
mechanism would also allow ongoing dialogue between 
data users and City officials, leading to more dynamic 
and responsive data management practices.

An especially valuable engagement avenue is through events such as NYC 
Open Data Week, which foster collaboration and innovation across sectors. 
Our involvement in this project stemmed from such an event, where we were 
inspired by the potential of open data to enhance public services. Governments 
should continue these types of initiatives, which not only promote 
data literacy, but encourage community-driven improvements to datasets. 
By leveraging the expertise of data scientists, statisticians, stakeholders, 
and civic organizations, governments can ensure their open datasets 
remain accurate, transparent, and useful for a wide range of applications.

@@

This research further emphasizes the necessity of harmonizing data
across City agencies, particularly resolving discrepancies in field
names, formats, usage, and definitions. Consistency in data is essential for
trustworthy machine learning, where inconsistencies across sources can
lead to distorted outcomes. This challenge is even more pronounced in
datasets that span long time periods, as changes in agency structures
or reporting standards which can introduce subtle biases and data 
distortions. Long-term data consistency is crucial for 
longitudinal studies and predictive modeling, as even minor 
data shifts can lead to significant deviations in model outcomes, 
as noted by \citet{rahm2000data} and\citet{borgman2012conundrum}.

@@

\section*{Supplementary Material}
The NYC 311 R code used to produce these results, along 
with additional analytical findings, are available 
at \url{https://github.com/jun-yan/nyc311clean}. Reference 
data files (311 SR data, USPS zip codes) are available 
at \url{https://doi.org/10.6084/m9.figshare.c.7617311}.

>>>>

\subsection{Structural Issues}
\label{sec:structural}
Structural issues refers to how data is organized, formatted, 
or presented within a dataset. Structural issues can cause difficulty in 
analyzing the data effectively, often necessitating extensive 
data cleaning. We identified three primary structural issues in the
NYC 311 dataset: undocumented fields, inconsistent representations of
missing data, and daylight saving time anomalies.

@@

\subsection{Accuracy and Precision}
\label{sec:precision}
The question of precision vs. accuracy arises with the \texttt{latitude} 
and \texttt{longitude} fields. as well as with the duplicative 
\texttt{location} field. These fields are expressed as
a 14-decimal number, e.g. \texttt{latitude} 40.86769186022511. 
Given that 1 degree of latitude at the equator is equal to 111.04 
kilometers, the ``1'' at the end of that decimal number represents 
approximately 1.1 nanometers (1/1,000,000,000 of a meter). (For 
reference a DNA molecule is approximately 2 nanometers in width.) This 
representation appears to be a classic case of extreme precision, albeit with limited 
accuracy. All the \texttt{latitude}, \texttt{longitude}, and \texttt{location}
fields display this characteristic.

@@

\section{Government Open Data Curation Recommendations}
\label{sec:recommendations}
Building upon  insights from our case study of the 
NYC 311 SR data, we propose a set of data 
curation principles tailored for government open datasets. 
These principles are designed to address the unique challenges 
 observed in the curation of such datasets, 
ensuring they remain reliable, consistent, and useful for 
public services, research, and other applications.
To facilitate implementation, the principles are prioritized based on
their impact and feasibility, offering a clear roadmap for improving
data curation practices.

\subsection{Principles}

\paragraph{Principle 1: Establishing Cross-Organizational Consistency}
A key challenge in government released datasets is the 
harmonization of data fields across various Agencies, and 
maintaining that consistency over time. Inconsistent field
definitions, data formats, or value domains can lead to erroneous
analyses and hinder cross-agency collaboration.
To address this, Agencies should adopt 
standardized naming conventions, value domains, usages, and 
formats across all datasets. Regular audits and data harmonization 
processes should be implemented to ensure consistency 
across long-term datasets, particularly as Agency structures 
and reporting standards evolve. For instance, ensuring that 
historical datasets reflect the same field naming conventions 
over a span of years can prevent inconsistent analyses. 
These standards should be promulgated by the Open Data 
governing authority, following agreements by the relevant parties.

\paragraph{Principle 2: Ensuring Data Entry Accuracy}
Data accuracy and validity are fundamental to the utility of 
open datasets. In the NYC 311 data, issues such as 
invalid zip codes and nonsensical date entries 
highlight the need for rigorous validation processes. Governments 
should establish clear protocols for data entry, ensuring that 
fields are populated with valid, legally acceptable values. Accurate
data entry reduces the risk of introducing systemic errors,
particularly in high-volume datasets. To
the maximum extent possible,the software should assist in 
ensuring data entries are accurate and valid. Some examples include:

@@

\paragraph{Principle 3: Optimizing Storage Efficiency}
Efficient storage and data representation are crucial for handling 
large government datasets. Our research found several duplicate and
near-duplicate fields in the 311 SR dataset. Removing these
redundancies could reduce the dataset size by 39.5\%, improving
performance without compromising analytical value. Eliminating 
redundant fields and encoding categorical variables can 
significantly reduce file sizes and improve the efficiency 
of data analysis. Similarly, categorical variables could be 
encoded in a standardized format to optimize storage and 
streamline analyses. Reducing file size not only saves 
storage space but also enhances the performance of queries, 
downloads,  and analysis processes.

\paragraph{Principle 4: Updating Data Dictionaries}
Clear and accurate data documentation is essential for the effective use 
of government datasets. Our study found several 
instances where the Data Dictionary did not accurately reflect the 
actual data. For example, undocumented fields and outdated domain
definitions in the 311 SR dataset introduced ambiguity, complicating
analysis. Domains of legal/allowable values should be referenced
in the Data Dictionary. Regular reviews and updates should 
occur whenever there are changes in data structure, data usage, or 
content. This practice enhances the ability of external users, 
including researchers and developers, to work 
effectively with the data.

\paragraph{Principle 5: Automating Quality Assurance Processes}
Automating data validation and quality assurance processes can 
significantly improve the reliability of government datasets. Certain
unusual trends can be identified, and irregularities highlighted. 
In the NYC 311 case, issues such as large spikes in SR 
creation and closure at exactly midnight (or noon) likely 
stem from a cron script process that distorts the 
accurate capture of SR durations. Real-time validation systems could
prevent such errors, ensuring data integrity at the point of entry. By
implementing
real-time and event-driven validation tools, governments 
can both prevent such errors from entering the dataset in the first 
place, as well as highlight anomalous trends. Automated 
systems can be designed to flag inconsistencies, incorrect values, 
illogical sequences, or illogical data entries. Automation can 
also facilitate faster detection and correction of errors, reducing 
the need for manual intervention and ensuring the data 
remains clean and reliable as it is continuously updated.

Transparency is a core tenet of open data, and governments must ensure that datasets are not only accurate but also easily accessible and readily understandable. **This includes publishing clear metadata, which enables users to understand the dataset’s structure, limitations, and intended use cases.** Additionally, governments should communicate transparently about data quality issues, such as notifying the public when errors in SR fields (e.g., erroneous \texttt{closed\_date(s)}) are discovered, including an explanation about how such issues will be resolved. By ensuring transparency and keeping the public informed, governments foster trust and promote wider use of datasets for research, civic engagement, and innovation.

\paragraph{Principle 6: Addressing Data Transparency}
Transparency is a core tenet of open data, and governments must ensure 
that datasets are not only accurate but also easily accessible and 
readily understandable. This involves providing clear 
metadata and making datasets available in user-friendly 
formats. This includes publishing clear metadata, which enables users
to understand the dataset’s structure, limitations, and intended use
cases. Additionally, governments should communicate transparently
about data quality issues, such as notifying the public when errors in SR 
fields (e.g., erroneous \texttt{closed\_ date(s)} are discovered, including
an explanation about how such issues will be resolved. By 
ensuring transparency and keeping the public informed, 
governments foster trust and promote wider use of datasets 
for research, civic engagement, and innovation.

\subsection{Actionable Steps}
While these principles are broad, actionable steps are required to 
implement them effectively. Prioritizing these steps will help
governments address the most critical data curation challenges
first. Governments can begin by investing in
real-time validation systems, regular updates to data dictionaries, 
and tools for encoding and standardizing data fields. This would include
systems that automate enforcement of data validation rules 
(e.g., flagging incorrect zip codes), significantly improving data 
quality. Automated systems can proactively detect and correct common
issues, reducing the manual workload.
Collaboration with city agencies and stakeholders is crucial
to ensure that data standards, formats, and metadata meet the needs 
of diverse users, including policymakers, researchers, and the 
general public. Regular communication between agencies and external 
users can help identify recurring issues and areas where improvements 
are needed. By facilitating regular communication and collaboration 
with various stakeholders, governments can ensure data remains relevant 
and actionable for public use.

Automated Quality Assurance techniques should be employed to 
monitor the quality of the data, as measured by agreed upon
standards, such as the \% of invalid zip codes. Such tools could
capture and display the quality of the data sets as presented
on appropriate dashboards. These dashboards provide a real-time
snapshot of data quality, making deviations immediately visible and
actionable. Deviations in quality would be
automatically highlighted, alerting appropriate corrective actions.

Engagement with data scientists and statisticians is key to improving 
data curation strategies. These experts can help design and refine 
curation processes, such as creating algorithms for automatic data 
cleaning, identifying inconsistent data entries, and ensuring the 
dataset(s) structure is maintained over time. Their involvement also
helps to establish best practices for validating incoming data and
maintaining consistency across data sources.
Feedback, such as the issues reported 
in this paper, benefit from a clear and accessible channel to the 311 SR data 
team and other relevant City Agencies. Establishing such feedback loops 
and other opportunities for public engagement, ensures that 
recommendations for improving data curation and validation 
are not only heard and discussed, but implemented. An iterative
feedback system enables timely resolution of issues and fosters
long-term improvements in data management.

An especially valuable engagement avenue is through events such as NYC 
Open Data Week, which foster collaboration and innovation across sectors. 
Our involvement in this project stemmed from such an event, where we were 
inspired by the potential of open data to enhance public services. Governments 
should continue these types of initiatives, which not only promote 
data literacy, but encourage community-driven improvements to
datasets. These events can strengthen public engagement and
demonstrate the societal value of open data initiative
By leveraging the expertise of data scientists, statisticians, stakeholders, 
and civic organizations, governments can ensure their open datasets 
remain accurate, transparent, and useful for a wide range of applications.

@@

While the findings here are specific to NYC 311 data, the principles 
proposed have broader applicability. Datasets in healthcare, 
transportation, and environmental monitoring can similarly benefit 
from these curation strategies, enabling improved outcomes across 
domains.
This research further emphasizes the necessity of harmonizing data
across City agencies, particularly resolving discrepancies in field
names, formats, usage, and definitions. Consistency in data is essential for
trustworthy machine learning, where inconsistencies across sources can
lead to distorted outcomes. This challenge is even more pronounced in
datasets that span long time periods, as changes in agency structures
or reporting standards which can introduce subtle biases and data 
distortions. Long-term data consistency is crucial for 
longitudinal studies and predictive modeling, as even minor 
data shifts can lead to significant deviations in model outcomes, 
as noted by \citet{rahm2000data} and\citet{borgman2012conundrum}.

@@

\section*{Supplementary Material}
The NYC 311 R code used to produce these results, along 
with additional analytical findings, are available 
at \url{https://github.com/jun-yan/nyc311clean}. Reference 
data files (311 SR data of 2022--2023, USPS zip codes) are available 
at \url{https://doi.org/10.6084/m9.figshare.c.7617311}.
